---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
The assignment makes use of the data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17568 observations in it.

## First of all, we need to set local settings to default for the C language that reflects North-American usage
```{r}
Sys.setlocale("LC_ALL", "C")
```

The assignment report will provide answers to the issues detailed below.


## Loading and preprocessing the data
1. Set the correct R working directory with the setwd() function.

2. For loading the data stored in the csv file in the working directory use the read.csv() function:
```{r}
# Clear the workspace
rm(list=ls())

# Load the raw activity data
activity_raw <- read.csv("activity.csv", stringsAsFactors=FALSE)
```

3. Process/transform the data, if necessary, into a format suitable for analysis:
```{r}
# Transform the date attribute to an actual date format
activity_raw$date <- as.POSIXct(activity_raw$date, format="%Y-%m-%d")

# Compute the weekdays from the date attribute
activity_raw <- data.frame(date=activity_raw$date, 
                           weekday=tolower(weekdays(activity_raw$date)), 
                           steps=activity_raw$steps, 
                           interval=activity_raw$interval)

# Compute the day type (weekend or weekday)
activity_raw <- cbind(activity_raw, 
                      daytype=ifelse(activity_raw$weekday == "saturday" | 
                                     activity_raw$weekday == "sunday", "weekend", 
                                     "weekday"))

# Create the final data.frame
activity <- data.frame(date=activity_raw$date, 
                       weekday=activity_raw$weekday, 
                       daytype=activity_raw$daytype, 
                       interval=activity_raw$interval,
                       steps=activity_raw$steps)

# Clear the workspace
rm(activity_raw)
```

To display the first few rows of the activity data frame:
```{r}
head(activity)
```


## What is the mean of the total number of steps taken per day?

For this part of the assignment, we ignore the missing values in the dataset.

1. Calculate the total number of steps taken each day:
```{r}
# Compute the total number of steps each day (NA values removed)
sum_data <- aggregate(activity$steps, by=list(activity$date), FUN=sum, na.rm=TRUE)

# Rename the attributes
names(sum_data) <- c("date", "total")
```

2. To display the total number of steps each day:
```{r}
# To display all days
sum_data

# Or to display the first few rows of the sum_data data frame
head(sum_data)
```

3. The histogram of the total number of steps taken each day is made using the following lines of code:
```{r}
hist(sum_data$total, 
     breaks=seq(from=0, to=25000, by=2500),
     col="red",
     xlab="Total Number of Steps", 
     ylim=c(0, 20),
     main="Histogram of the Total Number of Steps Taken each Day\n(NA Removed)")
```

4. Calculate and report the mean and median of the total number of steps taken per day:
```{r}
# The mean is:
mean(sum_data$total)

# The median is:
median(sum_data$total)
```
These formulas gives a mean and median of 9354.23 and 10395 respectively.

## What is the average daily activity pattern?

1. Making a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis):
```{r}
# Compute the means of steps accross all days for each interval
mean_data <- aggregate(activity$steps, 
                       by=list(activity$interval), 
                       FUN=mean, 
                       na.rm=TRUE)

# Rename the attributes
names(mean_data) <- c("interval", "mean")
```

To display the first few rows of the mean_data data frame:
```{r}
head(mean_data)
```

2. The following code creates the time series plot:
```{r}
# Compute the time series plot
plot(mean_data$interval,
        mean_data$mean,
        type="l",
        col="blue",
        lwd=2,
        xlab="Interval [minutes]",
        ylab="Average Number of Steps",
        main="Time-series of the Average Number of Steps per Intervals\n(NA Removed)")
```

3. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?:
```{r}
# We find the position of the maximum mean
max_pos <- which(mean_data$mean == max(mean_data$mean))

# We lookup the value of interval at this position
max_interval <- mean_data[max_pos, 1]

# The 5-minute interval that contains the maximum of steps, on average, across all days is
max_interval
```
The 5-minute interval that contains the maximum of steps, on average across all days, is 835.

## Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1. To calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NA's):
```{r}
# Clear the workspace
rm(max_interval)

# We use the trick that a TRUE boolean value is equivalent to 1 and a FALSE to 0.
NA_count <- sum(is.na(activity$steps))

# The number of NAs is:
NA_count
```
The number of NAs is 2304.

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc:
```{r}
# Clear the workspace
rm(NA_count)

# Find the NA positions
na_pos <- which(is.na(activity$steps))

# Create a vector of means
mean_vec <- rep(mean(activity$steps, na.rm=TRUE), times=length(na_pos))
```

The strategy used is to remplace each NA value by the mean of the steps attribute.

3. To create a new dataset that is equal to the original dataset but with the missing data filled in:
```{r}
# Replace the NAs by the means
activity[na_pos, "steps"] <- mean_vec

# Clear the workspace
rm(mean_vec, na_pos)
```

Displaying the first few rows of the new activity data frame:
```{r}
head(activity)
```

4. To make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?:
```{r}
# Compute the total number of steps each day (NA values removed)
sum_data <- aggregate(activity$steps, by=list(activity$date), FUN=sum)

# Rename the attributes
names(sum_data) <- c("date", "total")

# Compute the histogram of the total number of steps each day
hist(sum_data$total, 
     breaks=seq(from=0, to=25000, by=2500),
     col="orange", 
     xlab="Total Number of Steps", 
     ylim=c(0, 30), 
     main="Histogram of the Total Number of Steps taken Each Day\n(NA Replaced by Mean Value)")
```

The new mean and median are computed like this:
```{r}
# The new mean is:
mean(sum_data$total)

# The new median is:
median(sum_data$total)
```
These formulas gives a new mean and a new median of 10766.19 and 10766.19 respectively.

These values differ greatly from the estimates of the first part of the assignment. The impact of imputing the missing values is to have more data, hence to obtain a bigger mean and median value.

## Are there differences in activity patterns between weekdays and weekends?

1. To create a new factor variable in the dataset with two levels - “weekdays” and “weekend” indicating whether a given date is a weekday or weekend day:
```{r}
# The new factor variable "daytype" was already in the activity data frame
head(activity)
```

2. To make a panel plot containing a time series plot (i.e. type = "l") of the 5- minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis):
```{r}
# Clear the workspace
rm(sum_data)

# Load the lattice graphical library
library(lattice)

# Compute the average number of steps taken, averaged across all daytype variable
mean_data <- aggregate(activity$steps, 
                       by=list(activity$daytype, 
                               activity$weekday, activity$interval), mean)

# Rename the attributes
names(mean_data) <- c("daytype", "weekday", "interval", "mean")
```

To display the first few rows of the mean_data data frame:
```{r}
head(mean_data)
```

The time series plot take the following form:
```{r}
# Compute the time serie plot
xyplot(mean ~ interval | daytype, mean_data, 
       type="l", 
       lwd=1,
       col="purple",
       xlab="Interval", 
       ylab="Number of Steps", 
       layout=c(1,2))
```
